---
title: "Exploratry Match Data analysis"
author: "Sujith Prakash Parsa, Kavya Inampudi, Sathya Kiran Gandavarapu"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Setup

### Loading the required libraries

```{r}
library(tidyverse)
library(plotly)
library(ggplot2)
library(skimr)
```

## Loading the datasets 

```{r}
MatchResult <- read.csv("matches.csv")
```

```{r}
DeliveriesData <- read.csv("deliveries.csv")
```

## Understanding the data using skim() function

```{r}
skim(MatchResult)
```
```{r}
skim(DeliveriesData)
```

## ommiting the missing values from the data

```{r}
# umpire3 columns only consists of NA values it is redundant information
MatchResult$umpire3 <- NULL
head(MatchResult)
```

```{r}
# Removing the rows with no result
MatchResult <- MatchResult %>% filter(!(MatchResult$result == c("no result", "tie")))

```

```{r}
# Changing the name of a team because they are same
MatchResult[MatchResult == "Rising Pune Supergiant"] <- "Rising Pune Supergiants"
```


## Exploratory Data Analysis

### Finding the team with most wins

```{r}
MostWins <- MatchResult %>% count(winner)
MostWins <- MostWins %>% arrange(desc(n))
```

```{r}
MostWins %>% ggplot(aes(x = fct_reorder(winner, n), y = n, color = "black", fill = winner)) + 
  geom_col() +
  coord_flip() +
  labs(title = "Most wins by teams",
       x = "Teams",
       y = "Number of wins",
       caption = "IPL Dataset of matches from 2008 - 2017") +
  theme_bw() +
  theme()
  
```

### Teams who won the most when they won the toss

```{r}
MostToss <- MatchResult %>% filter(toss_winner == winner) %>% count(toss_winner)
```

```{r}
MostToss %>% ggplot(aes(x = fct_reorder(toss_winner, n), y = n, color = "black", fill = toss_winner)) + 
  geom_col() +
  coord_flip() +
  labs(title = "Most wins when a team won Toss",
       x = "Teams",
       y = "Number of wins",
       caption = "IPL Dataset of matches from 2008 - 2017") +
  theme_bw() +
  theme()
```

### Teams who chose to bat first and win

```{r}
Mostwins_Batfirst <- MatchResult %>% filter((toss_winner == winner) & (toss_decision == "bat")) %>% count(winner)
```

```{r}
Mostwins_Batfirst %>% ggplot(aes(x = fct_reorder(winner, n), y = n, color = "black", fill = winner)) + 
  geom_col() +
  coord_flip() +
  labs(title = "Most wins by teams when choose to Bat first",
       x = "Teams",
       y = "Number of wins",
       caption = "IPL Dataset of matches from 2008 - 2017") +
  theme_bw() +
  theme()
```

### Teams who chose to field first and win

```{r}
Mostwins_fieldfirst <- MatchResult %>% filter((toss_winner == winner) & (toss_decision == "field")) %>% count(winner)
```

```{r}
Mostwins_fieldfirst %>% ggplot(aes(x = fct_reorder(winner, n), y = n, color = "black", fill = winner)) + 
  geom_col() +
  coord_flip() +
  labs(title = "Most wins by teams when choose to Field first",
       x = "Teams",
       y = "Number of wins",
       caption = "IPL Dataset of matches from 2008 - 2017") +
  theme_bw() +
  theme()
```

### Players who won the most Player_of_match(MVP)

```{r}
Most_Player_of_match <- MatchResult %>% count(player_of_match) %>% filter(n > 10)
```

```{r}
Most_Player_of_match %>% ggplot(aes(x = fct_reorder(player_of_match, n), y = n, color = "black", fill = player_of_match)) +
  geom_col() +
  theme_bw() + 
  coord_flip() +
  labs(title = " Players who won the most Player_of_Matches(MVP)",
       x = "Name of the Player",
       y = " Number of MVPs",
       caption =  "IPL Dataset of matches from 2008 - 2017") +
  theme(legend.position = " ",)
```

### Distribution of teams winng by more than 50 runs

```{r}
MostWin_byruns <- MatchResult %>% filter(win_by_runs > 50) 
```

```{r}
MostWin_byruns %>% ggplot(aes(x = win_by_runs, color = "black", fill = winner)) +
  geom_histogram() +
  labs(title = "Distribution of teams winning by more than 50 runs",
       x = "Win by Runs",
       y = "Number of times the team won",
       caption = "IPL Dataset of matches from 2008 - 2017") +
  theme_bw() +
  theme()
```

### Similarly Distribution of teams win by more than 5 wickets

```{r}
MostWin_bywickets <- MatchResult %>% filter(win_by_wickets >= 5)
```

```{r}
MostWin_bywickets %>% ggplot(aes(x = win_by_wickets, color = "black", fill = winner)) +
  geom_histogram() +
  labs(title = "Distribution of teams winning by more than 5 wickets",
       x = "Win by wickets",
       y = "Number of times the team won",
       caption = "IPL Dataset of matches from 2008 - 2017") +
  theme_bw() +
  theme()
```

### Find how many unique matches have been recorded in deliveries data

```{r}
UniqueMatches <- unique(DeliveriesData$match_id)
UniqueMatches
```

This data set consists of data of 636 different matches including play-by-play data.


### Taking a look at each match individually

```{r}
Match1 <- DeliveriesData %>% filter((match_id == 1))
```

```{r}
Match1 %>% filter(inning == 1) %>% count(batsman_runs)
```
```{r}
Match1 %>% filter(inning == 1) %>% count(dismissal_kind)
```

```{r}

# This code is extracted from "https://plotly.com/r/pie-charts/"

data <- Winning_chase

fig <- plot_ly(data, labels = ~winner, values = ~n, type = 'pie')
fig <- fig %>% layout(title = 'Match won when Field first',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

fig

```




